Formalization of Forcing in Isabelle/ZF \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}

### 2020-01-25 IJCAR2020/absoluteness.tex
spelling

The concepts of relativization and absoluteness (due to Gödel, in his
proof of the relative consistency of $\AC$ \cite{godel-L}) 
are both
prerrequisites and powerful tools in working with transitive
models. A \emph{class} is simply a predicate $C(x)$ with at least one
free variable $x$.
The \emph{relativization} $\phi^C(\bx)$ of a set-theoretic
definition
$\phi$ (of a relation such
as ``$x$ is a subset of $y$'' or of a function like $y=\P(x)$) to
a class $C$ is obtained by restricting all of its quantifiers to $C$.

@@

The following schematic lemma synthesizes its internal version,%
\footnote{The use of such schematic goals and the orginal definition
  of the collection of lemmas \isatt{sep{\isacharunderscore}rules} are due to Paulson.}
\begin{isabelle}
\isacommand{schematic{\isacharunderscore}goal}\isamarkupfalse%
\ ZF{\isacharunderscore}pairing{\isacharunderscore}auto{\isacharcolon}\isanewline
\ \ \ \ {\isachardoublequoteopen}upair{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}A{\isacharparenright}\ {\isasymlongleftrightarrow}\ {\isacharparenleft}A{\isacharcomma}\ {\isacharbrackleft}{\isacharbrackright}\ {\isasymTurnstile}\ {\isacharquery}zfpair{\isacharparenright}{\isachardoublequoteclose}\isanewline
\isacommand{unfolding}\isamarkupfalse%
\ upair{\isacharunderscore}ax{\isacharunderscore}def\ \isanewline
\ \ \isacommand{by}\isamarkupfalse%
\ {\isacharparenleft}{\isacharparenleft}rule\ sep{\isacharunderscore}rules\ {\isacharbar}\ simp{\isacharparenright}{\isacharplus}{\isacharparenright}
\end{isabelle}
(the actual formula obtained is
\isatt{Forall(Forall(Exists(upair{\isacharunderscore}fm(2,1,0))))}) and
our \isatt{synthesize} method introduces a new term
\isatt{ZF{\isacharunderscore}pairing{\isacharunderscore}fm} for it.
\begin{isabelle}
\isacommand{synthesize}\isamarkupfalse%
\ {\isachardoublequoteopen}ZF{\isacharunderscore}pairing{\isacharunderscore}fm{\isachardoublequoteclose}\ \isakeyword{from{\isacharunderscore}schematic}\ {\isachardoublequoteopen}ZF{\isacharunderscore}pairing{\isacharunderscore}auto{\isachardoublequoteclose}%
\end{isabelle}

>>>>

The concepts of relativization and absoluteness (due to Gödel, in his
proof of the relative consistency of $\AC$ \cite{godel-L}) 
are both
prerequisites and powerful tools in working with transitive
models. A \emph{class} is simply a predicate $C(x)$ with at least one
free variable $x$.
The \emph{relativization} $\phi^C(\bx)$ of a set-theoretic
definition
$\phi$ (of a relation such
as ``$x$ is a subset of $y$'' or of a function like $y=\P(x)$) to
a class $C$ is obtained by restricting all of its quantifiers to $C$.

@@

The following schematic lemma synthesizes its internal version,%
\footnote{The use of such schematic goals and the original definition
  of the collection of lemmas \isatt{sep{\isacharunderscore}rules} are due to Paulson.}
\begin{isabelle}
\isacommand{schematic{\isacharunderscore}goal}\isamarkupfalse%
\ ZF{\isacharunderscore}pairing{\isacharunderscore}auto{\isacharcolon}\isanewline
\ \ \ \ {\isachardoublequoteopen}upair{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}A{\isacharparenright}\ {\isasymlongleftrightarrow}\ {\isacharparenleft}A{\isacharcomma}\ {\isacharbrackleft}{\isacharbrackright}\ {\isasymTurnstile}\ {\isacharquery}zfpair{\isacharparenright}{\isachardoublequoteclose}\isanewline
\isacommand{unfolding}\isamarkupfalse%
\ upair{\isacharunderscore}ax{\isacharunderscore}def\ \isanewline
\ \ \isacommand{by}\isamarkupfalse%
\ {\isacharparenleft}{\isacharparenleft}rule\ sep{\isacharunderscore}rules\ {\isacharbar}\ simp{\isacharparenright}{\isacharplus}{\isacharparenright}
\end{isabelle}
(the actual formula obtained is
\isatt{Forall(Forall(Exists(upair{\isacharunderscore}fm(2,1,0))))}) and
our \isatt{synthesize} method introduces a new term
\isatt{ZF{\isacharunderscore}pairing{\isacharunderscore}fm} for it.
\begin{isabelle}
\isacommand{synthesize}\isamarkupfalse%
\ {\isachardoublequoteopen}ZF{\isacharunderscore}pairing{\isacharunderscore}fm{\isachardoublequoteclose}\ \isakeyword{from{\isacharunderscore}schematic}\ {\isachardoublequoteopen}ZF{\isacharunderscore}pairing{\isacharunderscore}auto{\isachardoublequoteclose}%
\end{isabelle}

### 2020-01-25 IJCAR2020/conclusion.tex
spelling

The next task in our path is pretty clear: To develop the forcing
notions to obtain the independence of $\CH$ 
along with the prerrequisite combinatorial results, v.g.\ the
$\Delta$-system lemma. A development of cofinality is under way in a
joint work with E.~Pacheco Rodríguez, which is needed for a general
statement of the latter.

In a second release of \isatt{ZF-Constructible-Trans}, we intend to conform
it to the lines of \emph{Basic Set Theory (BST)}
\cite[I.3.1]{kunen2011set} in which elementary results have proofs
using alternatively Powerset or Replacement. The interest in this
arises because many natural set models (rank-initial segments of the
universe or the family $H(\kappa)$ of sets hereditarily of cardinality
less than $\kappa$) satisfy one of those axioms and not the
other. There are also still some older or less significant proofs
written in tactical (\textbf{apply}) format; we hope we will find the
time to translate them to Isar. Finally, the automation of formula
synthesis is on an early stage of development; finishing that module
will make writing our proofs of closure under various operations
faster, and also turn the set theory libraries  more
usable to other researchers.

>>>>

The next task in our path is pretty clear: To develop the forcing
notions to obtain the independence of $\CH$ 
along with the prerequisite combinatorial results, v.g.\ the
$\Delta$-system lemma. A development of cofinality is under way in a
joint work with E.~Pacheco Rodríguez, which is needed for a general
statement of the latter.

In a second release of \isatt{ZF-Constructible-Trans}, we intend to conform
it to the lines of \emph{Basic Set Theory (BST)}
\cite[I.3.1]{kunen2011set} in which elementary results have proofs
using alternatively Powerset or Replacement. The interest in this
arises because many natural set models (rank-initial segments of the
universe or the family $H(\kappa)$ of sets  of cardinality
less than $\kappa$ hereditarily) satisfy one of those axioms and not the
other. There are also still some older or less significant proofs
written in tactical (\textbf{apply}) format; we hope we will find the
time to translate them to Isar. Finally, the automation of formula
synthesis is on an early stage of development; finishing that module
will make writing our proofs of closure under various operations
faster, and also turn the set theory libraries  more
usable to other researchers.

### 2020-01-25 IJCAR2020/forces_definition.tex
spelling

It was not straightforward to use the recursion machinery of
Isabelle/ZF to define $\forceseq$ and $\forcesmem$. For this, we
defined a relation $\frecR$ on 4-tuples of elements of $M$, proved
that it is wellfounded and, more important, we also proved an
induction principle for this relation:
%
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ forces{\isacharunderscore}induction{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymAnd}{\isasymtau}\ {\isasymtheta}{\isachardot}\ {\isasymlbrakk}{\isasymAnd}{\isasymsigma}{\isachardot}\ {\isasymsigma}{\isasymin}domain{\isacharparenleft}{\isasymtheta}{\isacharparenright}\ {\isasymLongrightarrow}\ Q{\isacharparenleft}{\isasymtau}{\isacharcomma}{\isasymsigma}{\isacharparenright}{\isasymrbrakk}\ {\isasymLongrightarrow}\ R{\isacharparenleft}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isachardoublequoteclose}\footnotemark\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymAnd}{\isasymtau}\ {\isasymtheta}{\isachardot}\ {\isasymlbrakk}{\isasymAnd}{\isasymsigma}{\isachardot}\ {\isasymsigma}{\isasymin}domain{\isacharparenleft}{\isasymtau}{\isacharparenright}\ {\isasymunion}\ domain{\isacharparenleft}{\isasymtheta}{\isacharparenright}\ {\isasymLongrightarrow}\ R{\isacharparenleft}{\isasymsigma}{\isacharcomma}{\isasymtau}{\isacharparenright}\ {\isasymand}\ R{\isacharparenleft}{\isasymsigma}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isasymrbrakk}\isanewline
\ \ \ \ \ \  {\isasymLongrightarrow}\ Q{\isacharparenleft}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}Q{\isacharparenleft}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}\ {\isasymand}\ R{\isacharparenleft}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
\footnotetext{The logical primitives of \emph{Pure} are
\isatt{\isasymLongrightarrow}, \isatt{\&\&\&}, and \isatt{\isasymAnd}
(implication, conjunction, and universal
quantification, resp.), which operate on the meta-booleans
\isatt{prop}.}
%
and 
obtained both functions as cases of a another one, 
$\forcesat$, using a single recursion on $\frecR$. Then we obtained 
(\ref{eq:def-forcing-equality}) and (\ref{eq:def-forcing-membership})
as our corollaries \isatt{def{\isacharunderscore}forces{\isacharunderscore}eq} and
\isatt{def{\isacharunderscore}forces{\isacharunderscore}mem}.

@@

Another improvement over our previous report are a couple of new
technical results concerning recursive definitions: Paulson proved
absoluteness of functions defined by wellfounded recursion over a
transitive relation. Some of our definitions by recursion
(\emph{check} and \emph{forces}) do not fit in that scheme.  One can
replace the relation $R$ for its transitive closure $R^+$ in the
recursive definition because one can prove, in general, that
$F\!\upharpoonright\!(R^{-1}(x))(y) =
F\!\upharpoonright\!({R^+}^{-1}(x))(y)$ whenever $(x,y) \in R$.

>>>>

It was not straightforward to use the recursion machinery of
Isabelle/ZF to define $\forceseq$ and $\forcesmem$. For this, we
defined a relation $\frecR$ on 4-tuples of elements of $M$, proved
that it is well-founded and, more important, we also proved an
induction principle for this relation:
%
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ forces{\isacharunderscore}induction{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymAnd}{\isasymtau}\ {\isasymtheta}{\isachardot}\ {\isasymlbrakk}{\isasymAnd}{\isasymsigma}{\isachardot}\ {\isasymsigma}{\isasymin}domain{\isacharparenleft}{\isasymtheta}{\isacharparenright}\ {\isasymLongrightarrow}\ Q{\isacharparenleft}{\isasymtau}{\isacharcomma}{\isasymsigma}{\isacharparenright}{\isasymrbrakk}\ {\isasymLongrightarrow}\ R{\isacharparenleft}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isachardoublequoteclose}\footnotemark\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymAnd}{\isasymtau}\ {\isasymtheta}{\isachardot}\ {\isasymlbrakk}{\isasymAnd}{\isasymsigma}{\isachardot}\ {\isasymsigma}{\isasymin}domain{\isacharparenleft}{\isasymtau}{\isacharparenright}\ {\isasymunion}\ domain{\isacharparenleft}{\isasymtheta}{\isacharparenright}\ {\isasymLongrightarrow}\ R{\isacharparenleft}{\isasymsigma}{\isacharcomma}{\isasymtau}{\isacharparenright}\ {\isasymand}\ R{\isacharparenleft}{\isasymsigma}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isasymrbrakk}\isanewline
\ \ \ \ \ \  {\isasymLongrightarrow}\ Q{\isacharparenleft}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}Q{\isacharparenleft}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}\ {\isasymand}\ R{\isacharparenleft}{\isasymtau}{\isacharcomma}{\isasymtheta}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
\footnotetext{The logical primitives of \emph{Pure} are
\isatt{\isasymLongrightarrow}, \isatt{\&\&\&}, and \isatt{\isasymAnd}
(implication, conjunction, and universal
quantification, resp.), which operate on the meta-booleans
\isatt{prop}.}
%
and 
obtained both functions as cases of a another one, 
$\forcesat$, using a single recursion on $\frecR$. Then we obtained 
(\ref{eq:def-forcing-equality}) and (\ref{eq:def-forcing-membership})
as our corollaries \isatt{def{\isacharunderscore}forces{\isacharunderscore}eq} and
\isatt{def{\isacharunderscore}forces{\isacharunderscore}mem}.

@@

Another improvement over our previous report are a couple of new
technical results concerning recursive definitions: Paulson proved
absoluteness of functions defined by well-founded recursion over a
transitive relation. Some of our definitions by recursion
(\emph{check} and \emph{forces}) do not fit in that scheme.  One can
replace the relation $R$ for its transitive closure $R^+$ in the
recursive definition because one can prove, in general, that
$F\!\upharpoonright\!(R^{-1}(x))(y) =
F\!\upharpoonright\!({R^+}^{-1}(x))(y)$ whenever $(x,y) \in R$.

### 2020-01-25 IJCAR2020/forcing_theorems.tex
reediciones (faltan dos observaciones by Miguel)
spelling

As with the previous section, the proofs of the forcing theorems have two different
flavours: The ones for the atomic formulas proceed by using the
principle of 
\isatt{forces{\isacharunderscore}induction}, and then an induction on
$\formula$ wraps the former with the remaining cases (\isatt{Nand} and \isatt{Forall}).

@@

Examples of proofs  using the second kind of induction include
the basic \isatt{strengthening{\isacharunderscore}lemma} and the main
results in this section, the lemmas of  Density  (actually, its nontrivial
direction
\isatt{dense{\isacharunderscore}below{\isacharunderscore}imp{\isacharunderscore}forces})
and Truth, 
which we reproduce next.
%
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ density{\isacharunderscore}lemma{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}p{\isasymin}P{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}env{\isasymin}list{\isacharparenleft}M{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isasymle}length{\isacharparenleft}env{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}\ {\isasymlongleftrightarrow}\ dense{\isacharunderscore}below{\isacharparenleft}{\isacharbraceleft}q{\isasymin}P{\isachardot}\ {\isacharparenleft}q\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}{\isacharbraceright}{\isacharcomma}p{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ truth{\isacharunderscore}lemma{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ \isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}M{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\ \isanewline
\ \ \ \ \ {\isachardoublequoteopen}{\isasymAnd}env{\isachardot}\ env{\isasymin}list{\isacharparenleft}M{\isacharparenright}\ {\isasymLongrightarrow}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isasymle}length{\isacharparenleft}env{\isacharparenright}\ {\isasymLongrightarrow}\ \isanewline
\ \ \ \ \ \ {\isacharparenleft}{\isasymexists}p{\isasymin}G{\isachardot}\ {\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}{\isacharparenright}\ \ {\isasymlongleftrightarrow}\ \ M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}\ map{\isacharparenleft}val{\isacharparenleft}G{\isacharparenright}{\isacharcomma}env{\isacharparenright}\ {\isasymTurnstile}\ {\isasymphi}{\isachardoublequoteclose}
\end{isabelle}
%
From these results, the semantical characterization of the forcing
relation (the ``definition of $\forces$''
\cite[IV.2.22]{kunen2011set}) follows easily:
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ definition{\isacharunderscore}of{\isacharunderscore}forces{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}p{\isasymin}P{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}env{\isasymin}list{\isacharparenleft}M{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isasymle}length{\isacharparenleft}env{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}\ {\isasymlongleftrightarrow}\isanewline
\ \ \ \ \ {\isacharparenleft}{\isasymforall}G{\isachardot}\ M{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isasymand}\ p{\isasymin}G\ {\isasymlongrightarrow}\ M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}\ map{\isacharparenleft}val{\isacharparenleft}G{\isacharparenright}{\isacharcomma}env{\isacharparenright}\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}

>>>>

As with the previous section, the proofs of the forcing theorems have two different
flavors: The ones for the atomic formulas proceed by using the
principle of 
\isatt{forces{\isacharunderscore}induction}, and then an induction on
$\formula$ wraps the former with the remaining cases (\isatt{Nand} and \isatt{Forall}).

@@

Examples of proofs  using the second kind of induction include
the basic \isatt{strengthening{\isacharunderscore}lemma} and the main
results in this section, the lemmas of  Density  (actually, its nontrivial
direction
\isatt{dense{\isacharunderscore}below{\isacharunderscore}imp{\isacharunderscore}forces})
and Truth, 
which we state next.
%
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ density{\isacharunderscore}lemma{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}p{\isasymin}P{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}env{\isasymin}list{\isacharparenleft}M{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isasymle}length{\isacharparenleft}env{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}\ {\isasymlongleftrightarrow}\ dense{\isacharunderscore}below{\isacharparenleft}{\isacharbraceleft}q{\isasymin}P{\isachardot}\ {\isacharparenleft}q\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}{\isacharbraceright}{\isacharcomma}p{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ truth{\isacharunderscore}lemma{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ \isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}M{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\ \isanewline
\ \ \ \ \ {\isachardoublequoteopen}{\isasymAnd}env{\isachardot}\ env{\isasymin}list{\isacharparenleft}M{\isacharparenright}\ {\isasymLongrightarrow}\ arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isasymle}length{\isacharparenleft}env{\isacharparenright}\ {\isasymLongrightarrow}\ \isanewline
\ \ \ \ \ \ {\isacharparenleft}{\isasymexists}p{\isasymin}G{\isachardot}\ {\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}{\isacharparenright}\ \ {\isasymlongleftrightarrow}\ \ M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}\ map{\isacharparenleft}val{\isacharparenleft}G{\isacharparenright}{\isacharcomma}env{\isacharparenright}\ {\isasymTurnstile}\ {\isasymphi}{\isachardoublequoteclose}
\end{isabelle}
%
From these results, the semantical characterization of the forcing
relation (the ``definition of $\forces$''
\cite[IV.2.22]{kunen2011set}) follows easily:
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ definition{\isacharunderscore}of{\isacharunderscore}forces{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}p{\isasymin}P{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\ {\isachardoublequoteopen}env{\isasymin}list{\isacharparenleft}M{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\isasymphi}{\isacharparenright}{\isasymle}length{\isacharparenleft}env{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}\ {\isasymlongleftrightarrow}\isanewline
\ \ \ \ \ {\isacharparenleft}{\isasymforall}G{\isachardot}\ M{\isacharunderscore}generic{\isacharparenleft}G{\isacharparenright}{\isasymand}\ p{\isasymin}G\ {\isasymlongrightarrow}\ M{\isacharbrackleft}G{\isacharbrackright}{\isacharcomma}\ map{\isacharparenleft}val{\isacharparenleft}G{\isacharparenright}{\isacharcomma}env{\isacharparenright}\ {\isasymTurnstile}\ {\isasymphi}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}

### 2020-01-25 IJCAR2020/isabelle_metatheory.tex
menor: mudo comentarios de Pure, explico más main theorem
Arreglos muy pavos. Comentario sobre definiciones sintetizadas.
reediciones (faltan dos observaciones by Miguel)
spelling

Isabelle \cite{Isabelle,DBLP:books/sp/Paulson94} is a general proof
assistant based on fragment of higher-order logic called
\emph{Pure}. %% The logical primitives of this meta-level are
%% \isatt{\isasymLongrightarrow}, \isatt{\&\&\&}, and \isatt{\isasymAnd}
%% (implication, conjunction, and universal
%% quantification, resp.), which operate on the meta-booleans
%% \isatt{prop}. 
The results presented in this work are theorems of a
version of $\ZF$ set theory (without the Axiom of Choice, $\AC$) 
called \emph{Isabelle/ZF}, which is one of the
``object logics'' that can be defined on top of Pure (which is then
used as a language to define rules). Isabelle/ZF defines types
\isatt{i} and \isatt{o} for sets and booleans, resp., and the $\ZF$
axioms are written down as terms of type \isatt{o}.

@@

It should be noted that Pure is a very weak framework and has no
induction/recursion capabilities. So the only way to define functions
by recursion is inside the object logic. (This works the same for
Isabelle/HOL.) For this reason, to define the relation of forcing, we
needed to resort to \emph{internalized} first-order formulas: they
form a recursively defined set \isatt{formula}. For example, the
predicate of safisfaction
\isatt{sats::i{\isasymRightarrow}i{\isasymRightarrow}i{\isasymRightarrow}o}
(noted $M,\mathit{ms}\models\phi$ for a set $M$, for a set $M$,
$\mathit{ms}\in\isatt{list}(M)$ and $\phi\in\formula$
%% (noted \isatt{M, ms \isasymTurnstile{ }\isasymphi} for a set \isatt{M},
%% \isatt{ms {\isasymin} list(M)} and \isatt{{\isasymphi} {\isasymin} formula})
had already been defined by recursion in \cite{paulson_2003}. We added
the corresponding notion $M\models \Phi$ for satisfaction of a set of
sentences.

>>>>

Isabelle \cite{Isabelle,DBLP:books/sp/Paulson94} is a general proof
assistant based on fragment of higher-order logic called
\emph{Pure}. 
The results presented in this work are theorems of a
version of $\ZF$ set theory (without the Axiom of Choice, $\AC$) 
called \emph{Isabelle/ZF}, which is one of the
``object logics'' that can be defined on top of Pure (which is then
used as a language to define rules). Isabelle/ZF defines types
\isatt{i} and \isatt{o} for sets and booleans, resp., and the $\ZF$
axioms are written down as terms of type \isatt{o}.

@@

It should be noted that Pure is a very weak framework and has no
induction/recursion capabilities. So the only way to define functions
by recursion is inside the object logic. (This works the same for
Isabelle/HOL.) For this reason, to define the relation of forcing, we
needed to resort to \emph{internalized} first-order formulas: they
form a recursively defined set \isatt{formula}. For example, the
predicate of satisfaction
\isatt{sats::i{\isasymRightarrow}i{\isasymRightarrow}i{\isasymRightarrow}o}
(written $M,\mathit{ms}\models\phi$ for a set $M$,
$\mathit{ms}\in\isatt{list}(M)$ and $\phi\in\formula$)
%% (noted \isatt{M, ms \isasymTurnstile{ }\isasymphi} for a set \isatt{M},
%% \isatt{ms {\isasymin} list(M)} and \isatt{{\isasymphi} {\isasymin} formula})
had already been defined by recursion in \cite{paulson_2003}.

### 2020-01-25 IJCAR2020/replacement_choice.tex
spelling

Although our previous version of the developement used $\AC$, that was
only needed to show the Rasiowa-Sikorski Lemma (RSL) for
arbitrary posets. We have modularized the proof of the latter
and now the version for countable posets that we use to show the
existence of generic filters
does not require Choice (as it is well known). We also bundled the
full RSL along with our implementation of the principle of dependent
choices in an independent branch of the dependency graph, which is the
only place where the theory \texttt{ZF.AC} is invoked.

@@

\subsection{The main theorem}
With all these elements in place, we are able to transcript the main
theorem of our formalization:
\begin{isabelle}
\isacommand{theorem}\isamarkupfalse%
\ extensions{\isacharunderscore}of{\isacharunderscore}ctms{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ \isanewline
\ \ \ \ {\isachardoublequoteopen}M\ {\isasymapprox}\ nat{\isachardoublequoteclose}\ {\isachardoublequoteopen}Transset{\isacharparenleft}M{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ ZF{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\ \isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymexists}N{\isachardot}\ \isanewline
\ \ \ \ \ \ M\ {\isasymsubseteq}\ N\ {\isasymand}\ N\ {\isasymapprox}\ nat\ {\isasymand}\ Transset{\isacharparenleft}N{\isacharparenright}\ {\isasymand}\ N\ {\isasymTurnstile}\ ZF\ {\isasymand}\ M{\isasymnoteq}N\ {\isasymand}\isanewline
\ \ \ \ \ \ {\isacharparenleft}{\isasymforall}{\isasymalpha}{\isachardot}\ Ord{\isacharparenleft}{\isasymalpha}{\isacharparenright}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\isacharparenright}\ {\isasymand}\isanewline
\ \ \ \ \ \ {\isacharparenleft}M{\isacharcomma}\ {\isacharbrackleft}{\isacharbrackright}{\isasymTurnstile}\ AC\ {\isasymlongrightarrow}\ N\ {\isasymTurnstile}\ ZFC{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
Here, \isatt{\isasymapprox} stands for equipotence, \isatt{nat} is the
set of natural numbers, and the predicate 
\isatt{Transset} indicates transitivity; and as usual, \isatt{AC}
denotes the Axiom of Choice, and \isatt{ZF} and \isatt{ZFC} the
corresponding subsets of \isatt{formula}.

>>>>

Although our previous version of the development used $\AC$, that was
only needed to show the Rasiowa-Sikorski Lemma (RSL) for
arbitrary posets. We have modularized the proof of the latter
and now the version for countable posets that we use to show the
existence of generic filters
does not require Choice (as it is well known). We also bundled the
full RSL along with our implementation of the principle of dependent
choices in an independent branch of the dependency graph, which is the
only place where the theory \texttt{ZF.AC} is invoked.

@@

\subsection{The main theorem}
With all these elements in place, we are able to transcript the main
theorem of our formalization:
\begin{isabelle}
\isacommand{theorem}\isamarkupfalse%
\ extensions{\isacharunderscore}of{\isacharunderscore}ctms{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ \isanewline
\ \ \ \ {\isachardoublequoteopen}M\ {\isasymapprox}\ nat{\isachardoublequoteclose}\ {\isachardoublequoteopen}Transset{\isacharparenleft}M{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}M\ {\isasymTurnstile}\ ZF{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\ \isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymexists}N{\isachardot}\ \isanewline
\ \ \ \ \ \ M\ {\isasymsubseteq}\ N\ {\isasymand}\ N\ {\isasymapprox}\ nat\ {\isasymand}\ Transset{\isacharparenleft}N{\isacharparenright}\ {\isasymand}\ N\ {\isasymTurnstile}\ ZF\ {\isasymand}\ M{\isasymnoteq}N\ {\isasymand}\isanewline
\ \ \ \ \ \ {\isacharparenleft}{\isasymforall}{\isasymalpha}{\isachardot}\ Ord{\isacharparenleft}{\isasymalpha}{\isacharparenright}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\isasymalpha}\ {\isasymin}\ M\ {\isasymlongleftrightarrow}\ {\isasymalpha}\ {\isasymin}\ N{\isacharparenright}{\isacharparenright}\ {\isasymand}\isanewline
\ \ \ \ \ \ {\isacharparenleft}M{\isacharcomma}\ {\isacharbrackleft}{\isacharbrackright}{\isasymTurnstile}\ AC\ {\isasymlongrightarrow}\ N\ {\isasymTurnstile}\ ZFC{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
Here, \isatt{\isasymapprox} stands for equipotency, \isatt{nat} is the
set of natural numbers, and the predicate 
\isatt{Transset} indicates transitivity; and as usual, \isatt{AC}
denotes the Axiom of Choice, and \isatt{ZF} and \isatt{ZFC} the
corresponding subsets of \isatt{formula}.
